---
title: "BST.262 HW#2 ex1"
author: "C.V. Cosgriff"
output:
  html_document:
    df_print: paged
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

## Q1: Read data about opioid drug names into a dataframe
```{r}
opioid <- read_xlsx("./data/PartD_Prescriber_PUF_NPI_15_Drug_Category_Lists.xlsx", 
                    sheet = 2, range = "B3:B111")
```

## Q2: Load medicare data into spark cluster
```{r, message=FALSE, warning=FALSE}
# Load sparklyr and create cluster
library(sparklyr)

# Increase memory available to prevent crashing on the spark collect stage
# per the documentation.
Sys.setenv("SPARK_MEM" = "8g")
config <- spark_config()
config$spark.driver.memory <- "4G"
config$spark.executor.memory <- "4G"

sc <- spark_connect(master = "local", config = config)

partd_tbl <- spark_read_csv(sc = sc, name = "partd", 
               path = "./data/PartD_Prescriber_PUF_NPI_Drug_15.txt", 
               header = TRUE, delimiter = "\t")
```

## Q3: Calculate opioid prescribing rate by state

The goal is to calculate the opioid prescribing rate by state, where the rate is
defined as:

"opioid prescribing rate reflects the percentage of a prescriber’s total Part D claims that are opioid prescriptions"

From there, we simply write out the dataframe.

Without spark the code for this would look as follows:

```{r eval=FALSE}
partd %>% mutate(opioid = ifelse(drug_name %in% opioid$`Drug Name`, "opioid_claim", "non_opioid_claim")) %>%
  group_by(opioid, nppes_provider_state) %>% 
  summarise(total_claim_sum = sum(total_claim_count)) %>%
  spread("opioid", "total_claim_sum") %>%
  mutate(opioid_rate = opioid_claim / (opioid_claim + non_opioid_claim))
```

We can count on the dplyr verbs by using  sparklyR , but not the
tidyr verbs. Unfortunately there are multiple open pull requests regarding
implementing these in spark, (e.g. https://github.com/rstudio/sparklyr/issues/252),
with little movement since 2016.

To work around this, I'll use the opioid names to flag each claim as opioid
or not, as above, and then summarize by state and opioid; then I'll create
two dataframes for opioids and non-opioids, renaming the claim count columns.
From there, they can be joined, achieving the same result as tidyr::gather.

With that, the rate can be calculated, and the final dataframe can be written
to a CSV to complete Q3. 

```{r}
state_sum <- partd_tbl %>% 
  mutate(opioid = ifelse(drug_name %in% opioid$`Drug Name`, TRUE, FALSE)) %>% 
  group_by(nppes_provider_state, opioid) %>% 
  summarise(total_claim_sum = sum(total_claim_count))


opi <- state_sum %>% filter(opioid == TRUE) %>% mutate(total_opioid_claim = total_claim_sum) %>%
  select(nppes_provider_state, total_opioid_claim)
no_opi <- state_sum %>% filter(opioid == FALSE) %>% mutate(total_other_claim = total_claim_sum) %>%
  select(nppes_provider_state, total_other_claim)

state_sum <- inner_join(opi, no_opi, by = "nppes_provider_state") %>% 
  mutate(opioid_rate = (total_opioid_claim / (total_opioid_claim + total_other_claim)) * 100)

write.csv(state_sum, "state.opioidRate.csv")
```

The final result of Q3 is saved locally as state.opioidRate.csv.

Q4: Generate an interactive map with the data.
```{r}
# We use the leaflet library to generate a choriopleth
library(leaflet)

# My mapbox token for ccosgriff@hsph.harvard.edu account
mapbox_token <- "pk.eyJ1IjoiY2Nvc2dyaWZmIiwiYSI6ImNqYmhzeXQwajBsbmIyd29mY2xrcTlzM2EifQ.rlOgbpSddBAcyOMkVW7YhQ"

# Created map using the leafly guide from RStduio website; attained us-states.json
# from github.
states <- geojsonio::geojson_read("json/us-states.json", what = "sp")

# Load the opioid rate data we just calculated; to get it working with our map
# data requires we flip the abbreviations to state names, and to do this we need
# to add PR and DC to the built in lists state.abb and stata.name
state_sum <- read_csv("./state.opioidRate.csv") %>% 
  select(nppes_provider_state, opioid_rate)
names <- c(state.name, "Puerto Rico", "District of Columbia")
abb <- c(state.abb, "PR", "DC")
state_sum$nppes_provider_state <- names[match(state_sum$nppes_provider_state, abb)]
state_sum <- state_sum %>% filter(nppes_provider_state %in% states$name)

# Now we can get to generating the map
state_sum$opioid_rate <- round(state_sum$opioid_rate, 2)
bins <- round(quantile(state_sum$opioid_rate, probs = seq(0,1, by = (1/8)), 
         na.rm = TRUE), 2)

pal <- colorBin("YlOrRd", domain = state_sum$opioid_rate, bins = bins)

labels <- sprintf(
  "<strong>%s</strong><br/>Opioid Prescribing Rate: %g %%",
  states$name, state_sum$opioid_rate
) %>% lapply(htmltools::HTML)

leaflet(states) %>%
  setView(-96, 37.8, 4) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = mapbox_token)) %>%
  addPolygons(
    fillColor = ~pal(state_sum$opioid_rate),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
 addLegend(pal = pal, values = ~state_sum$opioid_rate, opacity = 0.7, title = "Opioid Prescribing Rate",
   position = "bottomright")
```

Screenshots of this map are in the local directory, and the map can be interacted
with by loading the knitted version of this file, also present in the local directory.

Q5: Retrieve the total population by state (you can use the get_decennial function) to calculate the Medicare
Part D opioid prescription rate per 10,000 people.
```{r}
library(tidycensus)
census_key <- "febef4def24f726dcbd616a62ebcaa324a07b448" # key I requested with hsph email

population <- get_decennial(geography = "state", variables = "H0100001",
                            year = 2010, key = census_key) %>%
  mutate(state_name = NAME) %>%
  select(state_name, value)

names <- c(state.name, "Puerto Rico", "District of Columbia")
abb <- c(state.abb, "PR", "DC")
population$state_name <- abb[match(population$state_name, names)]

partd_tbl <- partd_tbl %>% select(state_name = nppes_provider_state, drug_name,
                                  total_claim_count)

partd_wpop <- partd_tbl %>% inner_join(population, by = "state_name", copy = TRUE)

state_sum <- partd_wpop %>% 
  mutate(opioid = ifelse(drug_name %in% opioid$`Drug Name`, TRUE, FALSE)) %>% 
  filter(opioid == TRUE) %>%
  group_by(state_name) %>% 
  summarise(rate = sum(total_claim_count)/mean(value) * 10^4)

write.csv(state_sum, "state.opiodRatePer10k.csv")
```

The file that completes Q5 is saved as state.opioidRatePer10k.csv in the local
directory.